---
title: "Advanced: Building a module"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In [intermediate](https://antonellilab.github.io/outsider.devtools/articles/intermediate.html)
we demonstrated how a complex, command-line program with input and output files
(that we partly developed) can be cast as an `outsider` module.

On this page, we will develop a module for a typical command-line program with
lots of arguments.

## [`RAxML`](https://cme.h-its.org/exelixis/web/software/raxml/)

For this walkthrough we will create a module for "RAxML" a program
that can generate evolutionary trees using a maximum-likelihood. The exact
functioning (or use even!) of the program is not relevant, but it demonstrates
well the typical program which `outsider` aims to bring into the R environment:
lots of arguments, complex algorithm developed in an alternative language.


To demonstrate the program's complexity, here is the command's signature:

```
raxmlHPC[-SSE3|-AVX|-PTHREADS|-PTHREADS-SSE3|-PTHREADS-AVX|-HYBRID|-HYBRID-SSE3|HYBRID-AVX]
      -s sequenceFileName -n outputFileName -m substitutionModel
      [-a weightFileName] [-A secondaryStructureSubstModel]
      [-b bootstrapRandomNumberSeed] [-B wcCriterionThreshold]
      [-c numberOfCategories] [-C] [-d] [-D]
      [-e likelihoodEpsilon] [-E excludeFileName]
      [-f a|A|b|B|c|C|d|D|e|E|F|g|G|h|H|i|I|j|J|k|m|n|N|o|p|P|q|r|R|s|S|t|T|u|v|V|w|W|x|y] [-F]
      [-g groupingFileName] [-G placementThreshold] [-h] [-H]
      [-i initialRearrangementSetting] [-I autoFC|autoMR|autoMRE|autoMRE_IGN]
      [-j] [-J MR|MR_DROP|MRE|STRICT|STRICT_DROP|T_<PERCENT>] [-k] [-K] 
      [-L MR|MRE|T_<PERCENT>] [-M]
      [-o outGroupName1[,outGroupName2[,...]]][-O]
      [-p parsimonyRandomSeed] [-P proteinModel]
      [-q multipleModelFileName] [-r binaryConstraintTree]
      [-R binaryModelParamFile] [-S secondaryStructureFile] [-t userStartingTree]
      [-T numberOfThreads] [-u] [-U] [-v] [-V] [-w outputDirectory] [-W slidingWindowSize]
      [-x rapidBootstrapRandomNumberSeed] [-X] [-y] [-Y quartetGroupingFileName|ancestralSequenceCandidatesFileName]
      [-z multipleTreesFile] [-#|-N numberOfRuns|autoFC|autoMR|autoMRE|autoMRE_IGN]
      [--mesquite][--silent][--no-seq-check][--no-bfgs]
      [--asc-corr=stamatakis|felsenstein|lewis]
      [--flag-check][--auto-prot=ml|bic|aic|aicc]
      [--epa-keep-placements=number][--epa-accumulated-threshold=threshold]
      [--epa-prob-threshold=threshold]
      [--JC69][--K80][--HKY85]
      [--bootstop-perms=number]
      [--quartets-without-replacement]
      [---without-replacement]
      [--print-identical-sequences]
```

How are we going to code a module that is able to account for all these
arguments?

# Walkthrough

## Building

As before, let's just jump right in with the module skeleton.

```{r skeleton}
library(outsider.devtools)
flpth <- module_skeleton(repo_user = 'dombennett', program_name = 'raxml',
                         docker_user = 'dombennett', flpth = tempdir(),
                         service = 'github')
# folder name where module is stored
print(basename(flpth))
```

Next, the Dockerfile. This is a much more complicated installation process.
First we require the apt programs `wget`, `make` and `gcc` to download the
program and compile it. Then we need to download and compile the program.
The majority of the Dockerfile contents can be found in the installation
instructions of the
[RAxML GitHub site](https://github.com/stamatak/standard-RAxML#readme).

```{r dockerfile}
dockerfile_text <- "
FROM ubuntu:latest
RUN apt-get update && apt-get install -y \
    wget make gcc
# 8.2 for this demonstration
RUN wget https://github.com/stamatak/standard-RAxML/archive/v8.2.12.tar.gz && \
    tar zxvf v8.2.12.tar.gz && \
    rm v8.2.12.tar.gz && \
    mv standard-RAxML-8.2.12 raxml
RUN cd /raxml && make -f Makefile.SSE3.PTHREADS.gcc && \
    rm *.o && cp raxmlHPC-PTHREADS-SSE3 /usr/bin/.
RUN mkdir /working_dir
WORKDIR /working_dir
"
# write to latest Dockerfile
cat(dockerfile_text, file = file.path(flpth, 'inst', 'dockerfiles', 'latest',
                                      'Dockerfile'))
```

Dockerfile defined. What should our R function be?

```{r function-definition, eval=FALSE, include=TRUE}
#' @name raxml
#' @title raxml
#' @description Run raxml
#' @param arglist Arguments to raxml provided as a character vector
#' @param outdir Filepath to where all output files should be returned.
#' @example examples/example.R
#' @export
raxml <- function(arglist = arglist_get(...), outdir = getwd()) {
  files_to_send <- filestosend_get(arglist)
  arglist <- arglist_parse(arglist)
  otsdr <- outsider_init(pkgnm = 'om..raxml', cmd = 'raxmlHPC-PTHREADS-SSE3',
                         wd = outdir, files_to_send = files_to_send,
                         arglist = arglist)
  run(otsdr)
}
```

In this function, input and output files are intermixed in the arguments passed
to RAxML. To determine what is a file that should be sent to the container from
a normal argument, we have `filestosend_get`. This function goes through all the
provided arguments to check for any files (it tests whether they are valid file
paths) and returns them as a character vector to `files_to_send`. Then we have
`arglist_parse`, this converts all given arguments to their basename so that
absolute filepaths are dropped. All input files will be passed to the containers `working_dir/` and therefore no parental file path are required. This function
has additional functions for dropping irrelevant arguments in the context of an
`outsider` module (e.g. no need to specify an output directory at the 
command-line, this would need to be done at the initiation of the `outsider`
object.)

> **What's with this `outdir`-thingy?** The RAxML function returns a series of
files. To ensure they are all returned in a convenient location, we have
created an additional argument to the `raxml()` function called `outdir`. It can
be good practice to break-up information that needs to be processed by the
module
function into things like `arglist` and `outdir`. In some instances, you may
be calling a program where additional information can be provided outside of
remit of the main program being emulated. For example in our last example we had
`input_file` and `output_file`. Other examples include, number of threads,
memmory allocation for the program, etc.

OK! So that's what the function looks like. Let's write it to file.

```{r write-function-to-file}
function_text <- "
#' @name raxml
#' @title raxml
#' @description Run raxml
#' @param arglist Arguments to raxml provided as a character vector
#' @param outdir Filepath to where all output files should be returned.
#' @example examples/example.R
#' @export
raxml <- function(arglist = arglist_get(...), outdir = getwd()) {
  files_to_send <- filestosend_get(arglist)
  arglist <- arglist_parse(arglist)
  otsdr <- outsider_init(pkgnm = 'om..raxml', cmd = 'raxmlHPC-PTHREADS-SSE3',
                         wd = outdir, files_to_send = files_to_send,
                         arglist = arglist)
  run(otsdr)
}
"
# write to R/functions.R
cat(function_text, file = file.path(flpth, 'R', 'functions.R'))
```

Let's check the setup.

Ok, let's build, test and upload the module! (The test should pass because by
default, `-h` is used in the example and this is a valid argument for RAxML.)

```{r build-package}
module_build(flpth = flpth)
module_test(flpth = flpth)
```

And then upload ....

```{r upload, eval=FALSE, include=TRUE}
module_upload(flpth = flpth, code_sharing = TRUE, dockerhub = TRUE)
```

### Continuous Integration (GitHub only)

OK. So you have a module that works and passes its tests. How do we inform the
world that it works? The answer: [Travis-CI](https://travis-ci.org/)

Travis is a free service that monitors your GitHub for updates and then runs
a series of tests on a remote server to determine whether it passes its tests or
not. The result of these tests can be added to a repo's README in form of an
image badge: e.g. "Tests Passing".

To set-up travis for your repo, visit the webiste https://travis-ci.org/, sign
in with your GitHub account and then activate the service your repo of choice,
e.g. `om..raxml`.

Then we need to provide the test instructions for Travis in the form of a YAML
file called `.travis.yml` and upload this to our GitHub repo. The `.travis.yml`
can be created for `outsider` modules so:

```{r travis-file}
module_travis(flpth = flpth)
```

Then simply update the repo on GitHub!

## Dotting the Is and crossing the Ts

After you have built a module that passes its tests and is available for
download, there are a few extra things to consider:

* **Update the `om.yml`** Add a more detailed description
* **Update the README** The default README is very bland. It provides a template
for what sorts of things a good README should contain: installation, how to run,
where to get more help. Remember to only edit the `README.Rmd`. To update the
`README.md` for GitHub, you need to "knit" the `.Rmd` file by running
`devtools::build_readme()` -- this is the function `module_build()` runs.
* **Ensure discoverability** `outsider::module_search` can find `outsider`
module repos on GitHub if they have and `om.yml`, have name in the form `om..`
and have the phrase `outsider-module: ` in their description.

---

**Delete it all**

```{r clean-up}
module_uninstall(repo = 'om..raxml')
unlink(x = flpth, recursive = TRUE, force = TRUE)
```

---

# Tips and tricks

* The best way to learn how to build your own module is to look at how others
have created modules for programs you are familiar with. Check out the list of
already available modules at this
[page](https://antonellilab.github.io/outsider/articles/available.html).
* If you find issues with the docker image build step, you can try "logging
into" a container and running the installation steps to test out what works.
At the terminal, use
```{bash docker-build, eval=FALSE, include=TRUE}
# run an ubuntu container
docker run -t -d --name test ubuntu:latest
# iteractively run bash
docker exec -it test bash
# when finished, list all running containers and stop/rm
docker ps -a
docker stop test
docker rm test
```
