---
title: "Basic: Module Building"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`outsider` allows users to install and run programs on their own computer
without the need to leave the R environment. If there is a command-line program
that is not available through `outsider` you can create your own module!

If you are able to install the program on your own machine and have some
experience with R packages and GitHub then you should be readily able to
create one.


On this page, we will outline what an `outsider` module is and then provide
a simple walkthrough for creating the simplest of simple modules: a command line
program that prints whatever text the user provides it.

To follow this guide you will need the following:

* A GitHub account (sign-up [here](https://github.com/))
* A Docker-Hub account (sign-up [here](https://hub.docker.com/))

# Basics

## Module structure

At its heart, an `outsider` module is just an R package with a `Dockerfile`.
All modules have the following basic file/folder structure:

```
# Core files and folders of a module #
- DESCRIPTION
- R/
    - functions.R
    - import.R
- inst/
    - dockerfiles/
        - latest/
            - Dockerfile
    - om.yaml
- README.md
- README.Rmd
- .travis.yml
```

*R files and folders*

The R package is determined by the `DESCRIPTION` file and the `R/` folder.
The first of these describes the package details (package name, author,
dependencies, etc.) and the second contains the R code that makes up the
package.
The `R/` folder can have any number of scripts with whatever names a developer
chooses (skeleton default `functions.R` and `import.R`).

*Docker files and folders*

The `outsider` package and its modules depend on Docker to run. Docker acts
like a virtual machine on a user's computer. It allows users to run specific
code, often designed for different operating systems, on any machine that has
Docker installed. These code snippets are called Docker images. And these
images are described by a `Dockerfile`. The `dockerfiles/` folder in the
module contains the `Dockerfile` that describes how to install the external
command-line program that the develop wishes to run through R. `dockerfiles/`
can have multiple versions of a `Dockerfile` but every module must have a
`latest/`.

*GitHub files and folders*

To make modules discoverable on GitHub, all modules require an `om.yml`.
This file has two elements (program and details) encoded in the
[YAML format](https://en.wikipedia.org/wiki/YAML). In addition, the module
has a `README.md` file that provides the text describing the module on the
module's GitHub (or other code-sharing site) homepage. `.md` format is like a
simplified HTML. It can be generated by hand or -- by default with
`outsider.devtools` -- it can be rendered from a `.Rmd` file, in this case
`README.Rmd`. The advantage of the `.Rmd`-approach is that R code chunks are
parsed and their output is then included in the `.md` version of the file. This
allows users to better understand the code as they can directly see the output.
Finally, a `.travis.yml` provides instructions detailing how the module should
be tested on [Travis-CI](https://en.wikipedia.org/wiki/Travis_CI).

# Walkthrough

We will walk you through how to create your very own `outsider` module that will
simply print (through `echo`) any text provided. This process comes in a
series of steps: generate core files and folders, create Docker image,
document and build R package, try the module, upload to GitHub and Docker
Hub.

## Generate the files and folders

As displayed above, we need to generate the core files and folders of a module.
This process can be easily performed using the `module_skeleton()` function.
This function takes a few details about the developer and the program and then
generates all the necessary core file structures.

The necessary information required for the function to run are our GitHub and
Docker Hub usernames plus the name for the program we wish to provide as a
module (which is `echo` -- a UNIX command for printing). In these code snippets,
the usernames are of the `outsider` maintainer. In order for these examples to
work for you, you will need to change "dombennett" to your own usernames.
(Note, your usernames may differ for GitHub and Docker Hub).

```{r skeleton}
library(outsider.devtools)
flpth <- module_skeleton(repo_user = 'dombennett', program_name = 'echo',
                         docker_user = 'dombennett', flpth = tempdir(),
                         service = 'github')
# folder name where module is stored
print(basename(flpth))
```

The above code will create an `outsider` module with the name `om..echo`.
After running the above code, you should take a minute to inspect the generated
files. In particular you should look at the `DESCRIPTION` file, the `Dockerfile`
and `functions.R`.

> **Why `om..echo`?** All modules must start with "om.." in order for them to
be discovered on GitHub. This is not a requirement for the functioning of the
module, it just means `outsider::module_search()` can find them.

At this stage, we would then edit the `inst/dockerfiles/latest/Dockerfile` and
the `R/functions.R` to work for our chosen external program. But because `echo`
-- the program we wish to port through `outsider` -- is so simple we don't
actually have to make any changes to these files. By default, the `Dockerfile`
is based on Ubuntu which ships with `echo` and our starting function in
`functions.R` is based around running `echo`: it parses the arguments, creates
an outsider object, and then launches the object.

```{r echo-function, echo=TRUE, eval=FALSE}
# The echo function in om..echo
echo <- function(...) {
  # convert the ... into an argument list
  arglist <- arglist_get(...)
  # create an outsider object: describe the arguments and program
  otsdr <- outsider_init(pkgnm = 'om..echo',
                         cmd = 'echo', arglist = arglist)
  # run the command
  run(otsdr)
}
```

> **What's `...`?** In function calls in R, `...` indicate that any number of
arguments can be provided to a function. The `arglist_get` function will take
the `...` and convert them into a character vector that can be parsed.
`outsider` recommends module functions make use of this feature so that any
number of arguments can be passed to external programs. Additionally, the
developer would then not need to document all the arguments of the external
program. For many external programs there may be hundreds of arguments, all of 
which are likely to be already documented; viewable through commands like
`-h` or `--help`.

## Building the R package

With the skeleton set-up, we can now install the R package using `module_build()`.

```{r build-package}
module_build(flpth = flpth, build_documents = TRUE, build_package = TRUE,
             build_image = FALSE, build_readme = TRUE)
```

> **What does `build_documents = TRUE` do?** In addition to the core files
indicated above, an R package also requires R documentation files that are
stored in `man/` -- these provide the `?[function]` utility. The above
function generates these files via
[`roxygen`](https://github.com/klutometis/roxygen) comments and
tags located in the R scripts that make up the package, i.e. all comments that
begin `#'`. For more information, look up "object documentation in R".

## Creating/Building the Docker image

> **Dockerfile commands** Dockerfiles are series of instructions for
constructing a "containerised" machine called a Docker image (functionally it's
a bit like a virtual machine, but more lightweight). Each command begins
with a capitalised instruction followed by arguments. The most common
instruction would be "RUN", this executes command-line code within the Docker
image system. For example, `RUN echo "command!"` would pass "command!" to the
program "echo". All Dockerfiles begin with a FROM instruction, which pulls a
Docker image on which to build your own image. For example, many images are
built upon the Linux operating system Ubuntu in which case the first line of the
Dockerfile would be "FROM ubuntu:latest". This first line would then download
the latest Ubuntu Docker image and all subsequent "RUN" instructions would be
running in Ubuntu command-line. For far more detailed information on
Dockerfiles, see the
[Docker docs](https://docs.docker.com/engine/docker-overview/).

In our `om..echo` we have a dockerfiles folder that contains a Dockerfile
describing the Docker image for our `echo` program. Our "latest" Dockerfile
contains the instructions to pull the Docker image of the latest Ubuntu release,
to create a folder called "working_dir" and then set this new folder as the
"WORKDIR".

```
# Example host distro
FROM ubuntu:latest

# Install program using RUN lines

# outsider *requires* working_dir
RUN mkdir /working_dir
WORKDIR /working_dir
```

> **What's the WORKDIR?** The WORKDIR sets the starting directory when a command
is passed to the Docker image. All `outsider` modules require this to be
"working_dir" as it allows `outsider` functions to know where to transfer files
to and from the container/virtual machine.

Using our Dockerfile that has been preloaded with the installed R package, we
can now build our Docker image using `module_build()`. (In practice, we'd
combine steps 2 and 3 by calling `module_build` and setting all the build
arguments to TRUE.)

```{r docker-build}
module_build(flpth = flpth, tag = 'latest', build_image = TRUE,
             build_documents = FALSE, build_package = FALSE)
```

> **What is a tag?** A Docker tag is akin to the version number of the image.
By default, if no tag is provided,  Docker will use 'latest'.

> **Note**: The functions try to be helpful by printing to screen the exact
Docker commands of the tasks being performed. This is to give you a better idea
of what is happening and to allow you to recreate the operations via a terminal.

## Try the module.

With a Docker image and an installed R package we are now ready to try out the
module before we upload it online.

```{r test-package}
library(outsider)
# the repo always refers to the future github repo
echo <- module_import('echo', repo = 'dombennett/om..echo')
echo('hello world!')
```

## Upload to GitHub and Docker Hub

After we have played with the module and ensured it works as we would hope, we
can upload it to our GitHub and Docker accounts so that others may download it.

To upload to GitHub we must first create a version of the repository online.
Visit your GitHub account and then create a new repository by clicking
`Repositories > New`. Ensure to name the online version with the same name as
your module, i.e. `om..echo`. (No initial steps are required to upload to
Docker-Hub.)

```{R upload, eval=FALSE, include=TRUE}
# based on the module details, the function will determine which code-sharing
# service to upload to
module_upload(flpth = flpth, code_sharing = TRUE, dockerhub = TRUE)
```

---

**Delete it all**

Don't want `om..echo` on your computer? Delete it so ...

```{r clean-up}
# to delete the Docker image and uninstall the R package
module_uninstall(repo = 'dombennett/om..echo')
# to delete the repo folder
unlink(x = flpth, recursive = TRUE, force = TRUE)
```


---

## Next-up: [Intermediate](https://antonellilab.github.io/outsider.devtools/articles/intermediate.html)
