---
title: "Intermediate: Module Building"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In [basics](https://antonellilab.github.io/outsider.devtools/articles/intermediate.html)
we introduced the essential elements that make up a module and then built a
super simple module.

On this page, we will build a module for a slightly more complex program while
introducing some more functionalities of `outsider`.

## `figlet`

For this walkthrough, we will be creating a slightly more sophisticated module
than `om..echo` by building a module for [`figlet`](http://www.figlet.org/),
a program that takes textual input and returns ASCII art.

The program itself only takes text input and returns text output. We, however,
will use `outsider` to create a module that can either take an input file or
text and then return ASCII art to either the console or to an output file.

Not all command-line programs follow the same standards (different input/output
file arguments, different help pages ...), so `outsider` comes with a series of
functions to provide options for developers for how best they want to
encapsulate the program in module form.

(For reference, this module does already exist in the wild:
[dombennett/om..figlet](https://github.com/DomBennett/om..figlet))

# Walkthrough

## Building

To get started, we will first need our module skeleton.

```{r skeleton}
library(outsider.devtools)
flpth <- module_skeleton(repo_user = 'dombennett', program_name = 'figlet',
                         docker_user = 'dombennett', flpth = tempdir(),
                         service = 'github')
# folder name where module is stored
print(basename(flpth))
```

Next we will need to develop the Dockerfile! `figlet` is easy enough to install
on Ubuntu Linux, we can just use the "apt-get" system to install it. (Here I'm
editing the file from within R to make the process reproducible, you can edit
the file directly with a text editor.)

```{r dockerfile}
dockerfile_text <- "
# select Ubuntu Linux image
FROM ubuntu:latest
# Use the 'apt-get' system to install figlet
RUN apt-get update && apt-get install -y figlet
# set the working_dir/
RUN mkdir /working_dir
WORKDIR /working_dir
"
# write to latest Dockerfile
cat(dockerfile_text, file = file.path(flpth, 'inst', 'dockerfiles', 'latest',
                                      'Dockerfile'))
```

Now, we have our Dockerfile set-up and our module skeleton, we must develop the
R code. The skeleton code for our `figlet`-module is currently:

```{r skeleton-function, echo=TRUE, eval=FALSE}
#' @name figlet
#' @title figlet
#' @description Run figlet
#' @param ... Arguments
#' @example /examples/example.R
#' @export
figlet <- function(...) {
  # convert the ... into a argument list
  arglist <- arglist_get(...)
  # create an outsider object: describe the arguments and program
  otsdr <- outsider_init(pkgnm = 'om..figlet', cmd = 'figlet',
                         arglist = arglist)
  # run the command
  run(otsdr)
}
```

The above function would work fine. It would pass any arguments from the user
in R to the `figlet` program. But because `figlet` does not come with input/output arguments, we will need to hardcode these into our function and
use some if statements and bash to create this functionality. Because this will
require calling several commands via the terminal, we can create a shell script
that we can then pass through `outsider` to our container.

```{r rfunction}
function_text <- "
#' @name figlet
#' @title figlet
#' @description Run figlet with input and output files.
#' @param arglist Arguments for figlet
#' @param input_file Text input file
#' @param output_file ASCII art output file
#' @details If no input_file, will use arglist argument.
#' If no output_file, will print to console.
#' @example /examples/example.R
#' @export
figlet <- function(arglist = arglist_get(...), input_file = NULL,
                   output_file = NULL) {
  wd <- NULL
  # construct shell script from arglist
  arglist <- c('figlet', arglist)
  if (!is.null(input_file)) {
      # cat the input_file contents to figlet
      # (basename is used because on the container,
      #  filepaths cannot be used.)
      arglist <- c('cat', basename(input_file), '|', arglist)
  }
  if (!is.null(output_file)) {
      # write out the results of figlet to output_file
      arglist <- c(arglist, '>', basename(output_file))
      # determine where returned files should be sent
      wd <- dirpath_get(output_file)
  }
  # write arglist to temp file
  script <- file.path(tempdir(), 'script.sh')
  on.exit(file.remove(script))
  # ensure script is written in binary format
  script_cnntn <- file(script, 'wb')
  cmds <- paste(arglist, collapse = ' ')
  # debug print
  print(cmds)
  write(x = cmds, file = script_cnntn)
  close(script_cnntn)
  # initialise outsider container by specifying the command,
  # the arguments, the files to be sent, and the directory to where
  # returned files should be sent
  otsdr <- outsider_init(pkgnm = 'om..figlet', cmd = 'bash',
                         arglist = 'script.sh', wd = wd,
                         files_to_send = c(script, input_file))
  # run the command
  run(otsdr)
}
"
# write to R/functions.R
cat(function_text, file = file.path(flpth, 'R', 'functions.R'))
```

> **What does `arglist` do?** This function takes R objects and converts them
into character strings that then be passed to a command-line program. It can
take as many arguments as you want. For example, with `figlet` you may want to
convert the value of `a` to ASCII art. In this way, you would simply be able to
pass `a` to the `figlet` function rather than spell it out again.

> **What's with the `#'@`?** These are the roxygen tags. They determine the
documentation of the function. See
["Object documentation"](http://r-pkgs.had.co.nz/man.html) for more details.

Ok, let's build the module!

```{r build-package}
module_build(flpth = flpth)
```

Does it work?

```{r does-it-work}
library(outsider)
figlet <- module_import('figlet', repo = 'dombennett/om..figlet')
# without files
figlet(arglist = 'hello!')
# with input file
input_file <- file.path(tempdir(), 'input_figlet.txt')
cat('This is from the input text file!', file = input_file)
figlet(arglist = '', input_file = input_file)
# with output file
output_file <- file.path(tempdir(), 'output_figlet.txt')
figlet(arglist = 'Into the output file', output_file = output_file)
cat(readLines(con = output_file), sep = '\n')
# from input file to output file... with font block!
figlet(arglist = c('-f', 'block'), input_file = input_file,
       output_file = output_file)
cat(readLines(con = output_file), sep = '\n')
# clean-up
file.remove(c(input_file, output_file))
```

## Check, test and deploy

Ok! So we have a functioning module. But how do we know it's a functioning
module? To check our module structure, configuration and functioning we have a 
few helpful functions.

```{r check-and-test}
# make sure the folder structure is correct
module_check(flpth = flpth)
# check are the names of the module components
module_identities(flpth = flpth)
# test that the module works
# module_test(flpth = flpth)
```

Running `module_test` we will find that there is an issue with the module.
`module_test` checks the functioning of the code by running the example; our
current example tries to call `-h` with `figlet` -- which is
an invalid argument. To correct this we can simply updated our example.

```{r update-example}
example_text <- "
library(outsider)
figlet <- module_import('figlet', repo = 'dombennett/om..figlet')
figlet('hello!')
"
# write to examples/example.R
cat(example_text, file = file.path(flpth, 'examples', 'example.R'))
# re-build package
module_build(flpth = flpth, build_image = FALSE, verbose = FALSE)
```

Now re-running the test ...

```{r module_test}
module_test(flpth = flpth)
```


Nice! The package passes. That means we can upload the package to GitHub and
Docker-Hub using the same functions as we saw in "basic".

```{R upload, eval=FALSE, include=TRUE}
module_upload(flpth = flpth, code_sharing = TRUE, dockerhub = TRUE)
```

---

**Delete it all**

```{r clean-up}
module_uninstall(repo = 'om..figlet')
unlink(x = flpth, recursive = TRUE, force = TRUE)
```

---

## Next-up: [Advanced](https://antonellilab.github.io/outsider.devtools/articles/advanced.html)


